Elastic Scattering in General Relativistic Ray Tracing for Neutrinos

### 2017-07-11 paper.tex
Add Fig: cumulative f homogeneous slab

In the following we present two forms of the integrated distribution functions,
Eqns.~\ref{eqn:num_fae} and \ref{eqn:num_fse}:
1) the full form  by which the solution is $f_{\rm scat}=f_{\rm AE}+f_{\rm SE}$
(labeled `scat' since we include elastic scattering), and
2) the form treating only absorption/emission interactions
in which we set $\mathscr{K}_{\rm SE}=0$ so that $f_{\rm noscat}=f_{\rm AE}$
(labeled `no scat' since we ignore elastic scattering).
There is no simple relation between the two forms for a given ray; that is
$f_{\rm scat} \neq f_{\rm noscat}+f_{\rm SE}$.
This is because $f_{\rm AE}$ is integrated using solely the absorption optical
depth in the first form,
and the absorption plus scattering optical depth in the second form.
Before this work, the noscat form of the equations has been the standard for
ray tracing, though many authors include the scattering optical depth in
the integration of Eqn.~\ref{eqn:num_fae}
(see for example \cite{hari2010-gr_nunubar_collapsar}).
\todo{explain why; cite others}

\subsection{Infinite Homogeneous Slab:
  Testing Thermodynamic Equilibrium}
\label{ssec:test_equilibrium}
In an optically thick region the radiation field is in thermodynamic equilibrium
with the matter.
We set up a large slab of matter in two homogeneous configurations
representative of the fluid in the Massive Collapsed Star test
(Sec.~\ref{ssec:test_collapse}) at the radius 50~km.
For comparision with Collapsed Star test, we use the LS180 equation of state,
in which the equilibrium $\nu_e$ neutrino chemical potential is
$\mu_{\nu_e}=-0.575$~MeV
($\mu_{\bar{\nu}_e}=-\mu_{\nu_e}$ and $\mu_{\nu_x}=0$).
Plot in Fig.~\ref{fig:homogeneous_isotropic}.

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{figure-ferr-50km}
  \caption{Relative error in integrated equilibrium distribution functions
    in a homogeneous background at the thermodynamic point
    representative of the collapse profile at
    50~km ($\varrho=10^{11}\,{\rm g}\,{\rm cm}^{-3}$,
    $T=3.7$~MeV, $Y_e=0.12$),
    as described in Sec.~\ref{ssec:test_equilibrium}.
    As can be seen in Fig.~\ref{fig:mfps-50km} these three species
    over the given energies
    probe our numerical solution over length scales from
    100~m to $10^{5}\,{\rm km}$.
    The source of this error is the discarded boundary term
    $f_{\rm bdry}\sim10^{-6}$, described in Sec.~\ref{ssec:numerical}.
  }
  \label{fig:homogeneous_isotropic}
\end{figure}

>>>>

In the following we present two forms of the integrated distribution functions,
Eqns.~\ref{eqn:num_fae} and \ref{eqn:num_fse}:
1) the full form  by which the solution is $f_{\rm scat}=f_{\rm AE}+f_{\rm SE}$
(labeled `scat' since we include elastic scattering), and
2) the form treating only absorption/emission interactions
in which we set $\mathscr{K}_{\rm SE}=0$ so that $f_{\rm noscat}=f_{\rm AE}$
(labeled `no scat' since we ignore elastic scattering).
There is no simple relation between the two forms for a given ray; that is
$f_{\rm scat} \neq f_{\rm noscat}+f_{\rm SE}$.
This is because $f_{\rm AE}$ is integrated using solely the absorption optical
depth in the first form,
and the absorption plus scattering optical depth in the second form.
The differences between these two methods is apparent
in Fig.~\ref{fig:cumulative_f_homogeneous}.
Before this work, the noscat form of the equations has been the standard for
ray tracing, though many authors include the scattering optical depth in
the integration of Eqn.~\ref{eqn:num_fae}
(see for example \cite{hari2010-gr_nunubar_collapsar}).
\todo{explain why; cite others}

\subsection{Infinite Homogeneous Slab:
  Testing Thermodynamic Equilibrium}
\label{ssec:test_equilibrium}
In an optically thick region the radiation field is in thermodynamic equilibrium
with the matter.
We set up a large slab of matter in two homogeneous configurations
representative of the fluid in the Massive Collapsed Star test
(Sec.~\ref{ssec:test_collapse}) at the radius 50~km.
For comparision with Collapsed Star test, we use the LS180 equation of state,
in which the equilibrium $\nu_e$ neutrino chemical potential is
$\mu_{\nu_e}=-0.575$~MeV
($\mu_{\bar{\nu}_e}=-\mu_{\nu_e}$ and $\mu_{\nu_x}=0$).
Plot in Fig.~\ref{fig:homogeneous_isotropic}.
\todo{also describe this test with scat}

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{fig-cumulative_f-homogeneous}
  \caption{Cumulative distribution functions at $\varepsilon=11.25$~MeV
    in homogeneous infinite slab of Sec.~\ref{ssec:test_equilibrium}.
    The integration proceeds from left to right (backwards in time $t$).
    The two components of the scat solutions ($f_{\rm AE}$ in points,
    and $f_{\rm SE}$ in dashed lines) sum to equal the single component
    of the noscat solution ($f_{\rm AE}$ in points connected by lines).
    All solutions terminate when they achieve a total optical depth
    greater than $\tau_term=14$; in the scat cases this occurs at an
    earlier time since the total mean free paths are less than the
    absorption mean free paths (much less in the case of $\nu_x$).
    The positions of the points (only plotted for the $f_{\rm AE}$ components)
    are the time steps chosen by the adaptive time-stepping algorithm
    described in Sec.~\ref{ssec:numerical}.
  }
  \label{fig:cumulative_f_homogeneous}
\end{figure}

\begin{figure}
  \includegraphics[width=\columnwidth]{figure-ferr-50km}
  \caption{Relative error in integrated equilibrium distribution functions
    in the infinite homogeneous slab  described in
    Sec.~\ref{ssec:test_equilibrium}.
    Presented here are the relative differences between the final
    $f_{\rm AE}$ in the noscat case
    (see Fig.~\ref{fig:cumulative_f_homogeneous})
    and the equilibrium distribution functions given by
    Eqn.~\ref{eqn:feq}.
    As can be seen in Fig.~\ref{fig:mfps-50km} these three species
    over the given energies
    probe our numerical solution over length scales from
    100~m to $10^{5}\,{\rm km}$.
    The source of this error is the discarded boundary term
    $f_{\rm bdry}\sim10^{-6}$, described in Sec.~\ref{ssec:numerical}.
  }
  \label{fig:homogeneous_isotropic}
\end{figure}
